---
title: "exploratory_analysis"
author: "Dayoung Yu"
date: "3/24/2019"
output: pdf_document
---

```{r}
library(tidyverse)
library(kableExtra)
library(corrplot)
library(caret)

admit.data = read.csv(file = "./admission_predict.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    gre.std = (gre_score - mean(gre_score)) / sd(gre_score),
    toefl.std = (toefl_score - mean(toefl_score)) / sd(toefl_score),
    cgpa.std = (cgpa - mean(cgpa)) / sd(cgpa),
    uni.rating = university_rating,
    sop.strength = sop,
    lor.strength = lor
  ) %>% 
  select(gre.std:lor.strength, research, chance_of_admit)
```


### Explore correlations for GRE TOEFL CGPA
```{r}
cor.matrix = cbind(as.tibble(names(admit.data)), as.tibble(cor(admit.data)))

cor.matrix %>%
  kable("latex", caption = "Correlation matrix of predictors", digits = 4) %>%
  kable_styling(latex_options = c("hold_position","scale_down"))

x = model.matrix(chance_of_admit ~ ., admit.data)[,-1]
corrplot(cor(x))
```


From the correlation matrix, we see that all predictors except `research` are highly correlated with each other ($\rho$>0.5). The highest correlation occurs between `gre.std`, `toefl.std`, and `cgpa.std`, indicating that applicants who have high GPAs are more likely to have high test scores, and applicants who score well in one standardized test will more likely score well in the other. Including all three `gre.std`, `toefl.std`, and `cgpa.std` predictors may result in multicollinearity and highly variable model predictability. 


```{r}
x2 = model.matrix(chance_of_admit ~ gre.std+toefl.std+cgpa.std+sop.strength+lor.strength+uni.rating,
                  admit.data)[,-1]
y = admit.data$chance_of_admit
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)
featurePlot(x2, y, plot = "scatter", labels = c("","chance_of_admit"),
            type = c("p"), layout = c(3, 2))
```

The exploratory plots indicate the existence of a positive relationship between all predictors and `Chance of Admit`. From the second row, we see that as GPA, GRE, and TOEFL scores increase, the chance of admittance increases. In addition, higher ranks on a 1:5 scale of statement of purpose strength, letter of recommendation strength, and university rating are associated with high chances of admittance. The plot for `research` was not included due to its binomial nature. However, the data indicate that presence of research experience is also positively associated with chance of admittance.


```{r}
ggplot(data = admit.data, aes(x = gre.std, y = chance_of_admit)) + 
  geom_point()
ggplot(data = admit.data, aes(x = toefl.std, y = chance_of_admit)) + 
  geom_point()
ggplot(data = admit.data, aes(x = cgpa.std, y = chance_of_admit)) + 
  geom_point()
ggplot(data = admit.data, aes(x = sop.strength, y = chance_of_admit)) + 
  geom_point()
ggplot(data = admit.data, aes(x = uni.rating, y = chance_of_admit)) + 
  geom_point()
ggplot(data = admit.data, aes(x = lor.strength, y = chance_of_admit)) + 
  geom_point()
ggplot(data = admit.data, aes(x = research, y = chance_of_admit)) + 
  geom_point()
```


### SOP and LOR
```{r}
sop_plot = admit.data %>%
  ggplot(aes(x = sop.strength, y = chance_of_admit)) +
  geom_boxplot()

lor_plot = admit.data %>%
  ggplot(aes(x = lor.strength, y = chance_of_admit)) +
  geom_boxplot()

library(patchwork)
sop_plot + lor_plot
```

As expected, chance of admittance increases as strength of statement of purpose and letter of recommendation increases. We hypothesize that solely high SOP or LOR strength are not sufficient for admittance due to the fact there are many outliers of low chance_of_admit for high SOP/LOP strengths but only two outliers of high chance_of_admit for low SOP/LOP strengths.

### university rating
```{r}

admit.data %>%
  ggplot(aes(x = uni.rating, y = chance_of_admit)) +
  geom_boxplot()

```

Contrary to expectation, chance of admittance increases as university rating increases. One could expect that satisfactory ratings for a university would lead to higher competition, and thus lower the chances of admittance.

### research experience
```{r}
admit.data %>%
  mutate(research = as.factor(research)) %>%
  ggplot(aes(x = research, y = chance_of_admit)) +
  geom_boxplot()

```

Participants who have research experience have slightly higher changes of admittance than participants who do not have research experience.

