---
title: "linear_model"
author: "Justin Hsie"
date: "3/27/2019"
output: html_document
---

```{r cleaning, message = FALSE }
library(tidyverse)
admit.data = read.csv(file = "./admission_predict.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    gre.std = (gre_score - mean(gre_score)) / sd(gre_score),
    toefl.std = (toefl_score - mean(toefl_score)) / sd(toefl_score),
    cgpa.std = (cgpa - mean(cgpa)) / sd(cgpa),
    uni.rating = university_rating,
    sop.strength = sop,
    lor.strength = lor
  ) %>% 
  select(serial_no, gre.std:lor.strength, research, chance_of_admit)
```

#Creating the linear model

## What predictor variables did you include?

We are going to use all of the available predictors in the dataset. 

## What technique did you use, and why did you choose it? What assumptions, if any, are being made by using this technique?

We used glm

## If there were tuning parameters, how did you pick their values?

None

## How did you make your predictions?

10 fold cross validation. 

## Discuss the training/test performance if you have a test data set.

Same as above

## Which variables play important roles in predicting the response?


## What are the limitations of the models you used (if there are any)? Are the models flexible enough to capture the underlying truth?

```{r glm}
glm.fit = glm(chance_of_admit ~ gre.std + toefl.std + cgpa.std + uni.rating + 
                   sop.strength + lor.strength + research, data = admit.data)
plot(glm.fit)
```


